---
title: "Homework 1"
author: "Rong Ma"
date: "Sep 10 2015"
output: html_document
---

**
Bring rectangular data in
**

First we use read.delim and read.table function in R
```{r}
data1 = read.delim('http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt')
#data2 = read.table('http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt')
#Error in scan(file, what, nmax, sep, dec, quote, skip, nlines, na.strings,  : 
#  line 145 did not have 6 elements
```
From the R output, we can see the two functions give different results. On the one hand, read.delim gives a data frame with 1704 obs. of 6 variables. On the other hand, by using read.table, there is a report of error which shows "line 145 did not have 6 elements".

The reason for this difference is that, although read.delim is just a wrapper around read.table, the two functions have different settings of arguments such as "fill", "header", "sep" and so on.

To develop a call to read.table that is equivalent to read.delim, we should change some arguments as follows.
```{r}
data3 = read.table('http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt', header = T, fill = T, sep = "\t", quote = "\"", dec = ",", comment.char = "")
```
By ckecking the workspace, we find that data1 is the same as data3.

**
Smell test the data
**

We can try the following.
```{r}
is.data.frame(data1)
is.matrix(data1)
is.vector(data1)
is.list(data1)
```
And we know it's a data.frame.

As for the class and mode, we have
```{r}
mode(data1)
class(data1)
```

```{r}
str(data1)
```
So there are 6 variables and 1704 observations/rows.

Except for str, we also have summary, dim that can get these facts about “extent” or “size”.
```{r}
summary(data1)
dim(data1)
```
